---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Anciliary Function

I wrote a classifier for the time series indexes that will be used
to separate each day into `weekend`/`weekday` factor groups
```{r , echo=TRUE}
#Use custom function to apply type-of-day factor to a data frame
classify_week_day <-function(day_of_week) {
  if (day_of_week==6 | day_of_week == 0) {
    'Weekend'
  }
  else {
    'Weekday'
  }
}
```

## Loading and preprocessing the data

I took the original data and extracted a subset to use for the time series data that will be
needed for the aggregate calculations. I also went ahead and created aggregate vectors.

```{r,echo=TRUE}
#Load the necessary libraries
library(xts)
library(ggplot2)
library(lattice)

#Unzip the compressed data source
unzip('./activity.zip')

#Load the csv data from the activity data file
raw_step_data <- read.csv('./activity.csv',header=TRUE)

#Prep the date column so that it can be used for time series operations
raw_step_data$date <- as.Date(raw_step_data$date)

#Add a day-of-week column to use for coloring
raw_step_data$day_of_week <- as.POSIXlt(raw_step_data$date,tz="UTC")$wday

#combine the intervals with the dates for each step data item to prepare for time
#series calculations
raw_step_data$date <- as.POSIXct(raw_step_data$interval*60,origin=raw_step_data$date,tz="UTC")

#Extract a subset of the step data so that time-series manipulation can be performed
filtered_step_data <- subset(raw_step_data,!is.na(raw_step_data$steps))

#Build time series data structure which will be used for daily statistics (min,max,mean)
xts.ts <- xts(filtered_step_data$steps,filtered_step_data$date)

#Perform all step data aggregate calculations
total_steps_per_day <- apply.daily(xts.ts,sum)
average_steps_per_day <- apply.daily(xts.ts,mean)
min_steps_per_day <- apply.daily(xts.ts,min)
max_steps_per_day <- apply.daily(xts.ts,max)

#Construct new data frame that will hold all time-series aggregates
aggregate_step_data <- data.frame(cbind(seq(1,length(total_steps_per_day)),
                                  as.Date(index(total_steps_per_day)),
                                  as.numeric(total_steps_per_day),
                                  as.numeric(min_steps_per_day),
                                  as.numeric(max_steps_per_day),
                                  as.numeric(average_steps_per_day)))
#Add sensible column names
names(aggregate_step_data) <- c("day","date","daily_total","daily_min","daily_max","daily_average")
#Ensure date is in proper form for day-of-week population
aggregate_step_data$date <- as.Date(aggregate_step_data$date)
aggregate_step_data$day_of_week<-as.POSIXlt(as.Date(aggregate_step_data$day))$wday

```  
### Which data frame will be used for graphs and further exploratory analysis?  
I will use the newly constructed data frame , aggregate_step_data:

```{r, echo=TRUE}
   str(aggregate_step_data)  
   summary(aggregate_step_data)  
```



## What is mean total number of steps taken per day?

###Calculate the total number of steps taken per day
I stored the sum of steps per day in total_steps_per_day:
```{r,echo=TRUE}
  aggregate_step_data$daily_total
```  

###Make a histogram of the total number of steps taken each day
I leveraged the data from total_steps_per_day to generate the total steps per day histogram:
```{r, echo=TRUE}
  hist(aggregate_step_data$daily_total,breaks=50,
       xlab="Steps (per day)",main="Total Steps Taken Per Day",
       col=aggregate_step_data$day_of_week)
rug(aggregate_step_data$daily_total)
```  

###Calculate and report the mean and median of the total number of steps taken per day
I leveraged the data from aggregate_step_data to calculate the mean and mode for steps taken per
day. 
```{r,echo=TRUE}
  mean(aggregate_step_data$daily_total)  
  median(aggregate_step_data$daily_total)  
```


## What is the average daily activity pattern?

### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

I approached this question by taking the 61 days of averages , with each average coming from a particular day , and 
generated a time-series plot of that information:

```{r,echo=TRUE}
  plot(seq(1,59),aggregate_step_data$daily_average,
      type="l",xlab="Day",ylab="Average Steps Taken",
      main="Average Steps Taken Per Day")
```  

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

I approached this question by looking at the time-series plot from the previous graph and zeroing in on the day
that had the highest average. Judging by the graph , this looks to be an interval from day 52. I double-checked 
this against the actual average data:

```{r,echo=TRUE}
  aggregate_step_data$daily_average
  average_steps_per_day[52]
```

Next, I grabbed the actual time data from that day to locate the 5-minute interval that had the max value on that day:

```{r,echo=TRUE}  
  xts.ts['2012-11-23']
  max(xts.ts['2012-11-23'])
  
```

This led me to the 5-minute interval of : 2012-11-23 08:15:00  with a step-count of 567.  

## Imputing missing values

### Calculate and report the total number of missing values in the dataset

I calculated these numbers by doing the following, which takes the total and removes all rows without any NAs:

```{r,echo=TRUE}
  missing_data_count<-nrow(raw_step_data) - 
  sum(!is.na(raw_step_data$steps)&!is.na(raw_step_data$date)&!is.na(raw_step_data$interval))
  missing_data_count
```  


### Devise a strategy for filling in all of the missing values in the dataset.
### Create a new dataset that is equal to the original dataset but with the missing data filled in.

I will use a simple strategy that uses the global mean from the available data that is not missing any steps values:

```{r,echo=TRUE}
   global_average <- mean(filtered_step_data$steps)
   assign("imputed_step_data",raw_step_data)
   missing_step_idx <- is.na(imputed_step_data$steps)
   imputed_step_data[missing_step_idx,]$steps <- global_average
   
```



### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

I will perform these calculations using the same strategy as before , leveraging xts:

```{r,echo=TRUE}
#Build time series data structure which will be used for daily statistics (min,max,mean)
xts.imputed_ts <- xts(imputed_step_data$steps,imputed_step_data$date)
#Perform all step data aggregate calculations
imputed_total_steps_per_day <- apply.daily(xts.imputed_ts,sum)
imputed_average_steps_per_day <- apply.daily(xts.imputed_ts,mean)
imputed_min_steps_per_day <- apply.daily(xts.imputed_ts,min)
imputed_max_steps_per_day <- apply.daily(xts.imputed_ts,max)

#Construct new data frame that will hold all time-series aggregates
imputed_aggregate_step_data <- data.frame(cbind(seq(1,length(imputed_total_steps_per_day)),
                                          as.Date(index(imputed_total_steps_per_day)),
                                          as.numeric(imputed_total_steps_per_day),
                                          as.numeric(imputed_min_steps_per_day),
                                          as.numeric(imputed_max_steps_per_day),
                                          as.numeric(imputed_average_steps_per_day)))
#Add sensible column names
names(imputed_aggregate_step_data) <- c("day","date","daily_total","daily_min","daily_max","daily_average")
#Ensure date is in proper form for day-of-week population
imputed_aggregate_step_data$date <- as.Date(imputed_aggregate_step_data$date)
imputed_aggregate_step_data$day_of_week<-as.POSIXlt(as.Date(imputed_aggregate_step_data$day))$wday
```

### Do these values differ from the estimates from the first part of the assignment? 

I will use a simple summary statistic comparison:

```{r,echo=TRUE}
  summary(aggregate_step_data)
  summary(imputed_aggregate_step_data)
```  

I did not see any real difference. I would suspect that this is because the aggregates per day would be expected to stay relatively stable, since no matter which strategy is used , there was much more data that was originally used than the amount which was added after we fixed the missing values.


### What is the impact of imputing missing data on the estimates of the total daily number of steps?

I did not see any real impact due to imputing of the missing data.


## Are there differences in activity patterns between weekdays and weekends?
First, I needed to add the 'Weekend'/'Weekday' factor to the imputed aggregate data set:
```{r,echo=TRUE}
  imputed_aggregate_step_data$type_of_day<-mapply(classify_week_day,imputed_aggregate_step_data$day_of_week)
```
To answer this question, I graphed the data with the 'Weekend'/'Weekday' factor being used:

```{r,echo=TRUE}
  xyplot(imputed_aggregate_step_data$daily_average~seq(1:59)|factor(imputed_aggregate_step_data$type_of_day),
         type='l',layout=c(1,2),
         xlab='Interval',ylab='Number of Steps')
```

The comparison between activity on the weekend vs during the weekdays seems to show the weekdays as having more activity.
